### 2020-12-02 paper/sex_selection_spacing-ver5.tex
Initial commit of version 5 of paper
Draft version of conceptual framework

\author{Claus C P\"ortner\\
    Department of Economics\\
    Albers School of Business and Economics\\
    Seattle University, P.O. Box 222000\\
    Seattle, WA 98122\\
    \href{mailto:cportner@seattleu.edu}{\texttt{cportner@seattleu.edu}}\\
    \href{mailto:claus@clausportner.com}{\texttt{claus@clausportner.com}}\\
    \href{http://www.clausportner.com}{\texttt{www.clausportner.com}}\\
    \& \\
    Center for Studies in Demography and Ecology \\
    University of Washington\\ \vspace{2cm}
    }

\date{August 2020}

@@

\section{Background and Conceptual Framework}

To set the stage for the subsequent analyses and provide a conceptual framework for 
understanding birth spacing, I first discuss female education and labor force 
participation in India and relevant theories on birth spacing.

% [Female education]

Female education is a crucial explanatory variable here for three reasons.
First, higher female education is associated with lower fertility and increased use of 
sex selection
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Second, female labor force participation in India, as in other developing countries, 
first decreases and then increases with education
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
Finally, child mortality decreases with maternal education
\citep{Rosenzweig1982a,Whitworth2002,Maitra2008}.

% [TK where to put son preference discussion?]
% %
% \footnote{
% It is also possible that the effect is driven by a stronger son preference, although 
% higher education women have shown declining son preference \citep{bhat03,pande07}.
% }

I divide education levels into four groups: No education, 1--7 years, 8--11 years, and 
12 and more years.
The latter two correspond to having completed primary and secondary school, 
respectively.%
\footnote{
Although there are variations by state, elementary education in India consists of a 
primary school covering grades one through five and an upper primary---or middle 
school---covering grades six through eight. 
Similarly, secondary education covers grades nine and tenth for ``secondary education'' 
and 11 and 12 for ``upper secondary.''
}
To ensure that the results are comparable with the prior literature on fertility
and mortality in India, I follow the NFHS reports, except that I combine the 
less than five years and 5--7 years of schooling completed and the 8--9 
and 10--11 years of schooling completed.
This grouping allows me to capture the differences across education levels discussed below, 
while also having groups large enough for the empirical method.

% [what has happened to education]
Female education has increased substantially over time in rural and urban areas but 
is still substantially higher in urban than in rural areas.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 
20 years or older, whether married or not, based on the four rounds of the NFHS.
In rural areas, women with no education have gone from 90\% for the 1930s cohorts to 
less than 20\% for the 1990s cohorts.
Women with eight or more years of education have gone from almost zero for the 1930 cohort 
to more than 60\% for the 1990s cohorts, with about half in the 8--11 group and the 
other half in the 12 plus group.
In urban areas, the proportions with no education or 1--7 years have each declined 
to just below 10\%.
Most of the increase in urban female education came from the 12 plus group, which 
now accounts for more than half of all urban women for the most recent cohorts.

@@

% [female labor force participation]

The standard economic argument for shorter spacing with increasing female education 
is that parents incur time costs when they have children \citep{Hotz1997,schultz97}.
Suppose having children requires the mother to reduce her market work. 
Parents can then lower this cost of children by shortening birth spacing to take advantage 
of economies of scale in childrearing  \citep{Vijverberg1982}.

However, even as the level of female education has increased, the female labor force 
participation in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
India's female labor force participation is now lower than that of most other countries and does 
not yet show any signs of increasing \citep{Klasen2015,Chatterjee2018}.
In line with previous research, the NFHS data show a U-shaped relationship between 
education and working for married women, with the highest percentage working being women with 
either no education or 12 or more years and the lowest being women with 8--11 years of education.%
\footnote{
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

@@

% [sex selection]

The introduction of sex selection allows parents to avoid giving birth to girls but 
increases the expected interval to the next birth.
Theory suggests that sex selection increases with lower desired fertility and with
higher parity for a given desired number of children \citep{Portner2015b}.
Sex selection is more widespread among better-educated than less-educated and among urban 
than rural women, which is consistent with lower desired fertility increasing the
use of sex selection 
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated and urban women also tend to live in households with higher income, which 
lowers the relative costs of using sex selection and having long birth intervals.

The combination of rising incomes and continued son preference may lead to even longer 
spacing than might be expected from the income effect alone.
As women's education increases, their productivity in the production of offspring human 
capital also increases.
With relatively more boys born because of increased access to sex-selective 
abortions and the increasing income potential for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.

If more and ``better'' parental attention per child results in higher child ``quality,'' 
we should expect longer birth intervals \citep{Zajonc1975,Zajonc1976,Razin1980}.
However, the evidence on spacing's effect on child quality measures such as IQ 
and education is mixed for developed countries and nonexisting for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is health and mortality, where longer spacing does lead to better outcomes, 
although this relationship weakens with maternal education 
\citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

% If parents realize that longer spacing leads to better child outcomes, we may even observe 
% later conception, especially when, with sex selection, the next child is likely to be a boy.
% Working in the opposite direction of this "delay" effect is that women with more education 
% can, in principle, space their children closer together without substantially increasing 
% child mortality risk. 
% However, given the apparent lack of pressure to return to the labor force, this effect is 
% likely small.

>>>>

\author{Claus C P\"ortner\\
    Department of Economics\\
    Albers School of Business and Economics\\
    Seattle University, P.O. Box 222000\\
    Seattle, WA 98122\\
    \href{mailto:cportner@seattleu.edu}{\texttt{cportner@seattleu.edu}}\\
    \href{http://www.clausportner.com}{\texttt{www.clausportner.com}}\\
    \& \\
    Center for Studies in Demography and Ecology \\
    University of Washington\\ \vspace{2cm}
    }

\date{December 2020}

@@

\section{Conceptual Framework}

This section provides a conceptual framework for understanding how birth spacing changes
and guiding the subsequent analysis.
In non-Western countries, lower fertility is associated with longer birth spacing, but 
outside a ``natural fertility'' regime it is not clear that lower fertility necessarily 
leads to longer spacing \citep{Newman1988,Rutstein2011,Casterline2016}.
Thus, I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I argue that female education and area of residence are the most important factors 
to consider in the empirical analysis and discuss how they tie into the three theories.

The dramatic improvements in economic conditions in India is best illustrated by the 
doubling of real wages between 1987 and 2011 \citep{Klasen2015}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as a good becomes more expensive relative to other
goods, people substitute away from the now more expensive good and towards other goods.
With higher wages, the substitute effect predicts that people work more and spend less 
time on other time-intensive non-wage earning activities, such as children or leisure.
The income effect captures that even if we ignore the substitution effect and hold hours 
worked constant higher wages increase the available income. 
This positive income effect leads people to spend more time on time-intensive activities 
such as children and leisure and less time working.

Empirically, higher female wages generally lowers fertility, whereas higher male 
wages increases fertility \citep{Hotz1997,schultz97}.
This pattern is consistent with women spending substantially more time on child care 
than men, leading to the substitution effect dominating for women while the income
effect dominates for men.

In addition to lowering fertility, higher female wages will tend to shorten spacing between 
children if having children requires the mother to reduce her economic activities.
Specifically, shortening birth spacing allows parents to take advantage of economies of 
scale in childrearing---for example, looking after two children simultaneously requires 
less than double the time needed to look after one child \citep{Vijverberg1982,Hotz1997}.
The reduction in the length of birth intervals from higher female wages will be more 
pronounced the more women participate in the labor force rather than working at home as 
economic activities at home are more compatible with child rearing.

Parents care not only about how many children they have but also about the outcomes
for the children, and, everything else equal, more children means less resources to invest 
per child.
Hence, with declining fertility we should see better child outcomes such as lower 
mortality and more schooling.
However, with increasing income, it is, in principle, possible to have both more
children and better outcomes for them.
Increasing returns to education might make the parents reduce fertility and invest 
more in each of the children's education \citep{Rosenzweig1982a}.

Longer spacing between births is generally expected to have positive effects on 
child outcomes and if fertility falls because of increased desired for better child
outcomes we would, therefore, also expect spacing to increase.
The clearest example of birth spacing affecting child outcomes is for health,
where very short spacing---approximately 24 months or less---leads to worse health 
and mortality outcomes for children 
\citep{Whitworth2002,Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.
More speculative is whether the increased individual parental attention per child that 
should arise from longer spacing results in better human capital outcome for children 
\citep{Zajonc1975,Zajonc1976,Razin1980}.
The evidence on spacing's effect on child quality measures such as IQ and education is 
mixed for developed countries and nonexisting for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

Empirically, son preference has only a minor effect on overall fertility, but does impact 
the sex composition within families through differential stopping behavior 
\citep{repetto72,leung94,clark00,Basu2010,Barcellos2014}. 
Furthermore, as mentioned above, stronger son preference leads to shorter spacing and the 
associated worse health outcomes for daughters when there are no sons in the household 
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.

The use of sex selection allows parents to avoid giving birth to a girl but each
abortion increases the expected interval to the next birth by 6--12 months.
This increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation.

Holding son preference constant, lower desired fertility increases the use of sex 
selection, suggesting that the first order effect of higher female wages is to increase 
sex selection because of the fertility effect discussed above 
\citep{Portner2015b,Jayachandran2017}.  
Furthermore, the higher the parity the more likely the use of sex selection if the 
desired number of sons has not been reached.
[cost of sex selection?]

% Factors/predictions
Female education and area of residence are the two principal explanatory variables I use
in the empirical analyses.
The main reason is that female education and area of residence are the closest proxies in 
the data for wages and access to and cost of sex selection, none of which are available.
In addition, both higher female education and urbanization are associated with lower 
fertility and increased use of sex selection in past research
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.

Finally, India has seen tremendous increases in both education and urbanization.
The proportion of the population living in urban areas almost doubled from 18\%
in 1960 to 35\% in 2019 \citep{United-Nations2019}.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 20 years or older, whether married or not, in the 
four rounds of the NFHS.
Female education has increased substantially in both rural and urban areas,
but is still considerably higher in urban than in rural areas.
In rural areas, women with eight or more years of education have gone from almost zero for 
the 1930 cohort to more than 60\% for the 1990s cohorts, with about half in the 8--11 
group and the other half in the 12 plus group.
In urban areas, most of the increase in urban female education came from the 12 plus 
group, which now accounts for more than half of all urban women for the most recent cohorts.

@@

Families in urban areas are likely to have fewer children and higher rates of sex 
selection and, therefore, increasingly longer birth intervals over time compared to
rural families.
With the costs of children higher in urban than in rural areas both because of the
lower opportunities for productive work of children and the higher cost of consumption
related to children, fertility is expected to be lower in urban than in rural areas.
Furthermore, access to health care and generally a healthier environment because of,
for example, cleaner water means that urban parents can have fewer children and invest
more in each than in rural areas.%
\footnote{
However, holding parental education constant, there is evidence that children in
Indian slums do worse than they would have in rural areas \citep{Portner2018a}.
}
However, the better access to health care also means that there is easier access to
sex selection and the generally higher income of urban households makes it more
affordable.
[differences in birth intervals without sex selection?]

Wages generally increase with education suggesting that higher female education
should shorten birth spacing and lower fertility, but also increase the use of sex 
selection.
However, female education undoubtedly represents more than the value of women's time.
Two examples are the productions of child health and offspring human capital.
[something on norms?]

First, children of more educated women tend to have better health outcomes and lower 
mortality than less educated women, either from higher income or better health knowledge 
\citep{Rosenzweig1982a,Kovsted2002,Whitworth2002,Maitra2008}.
Lower expected mortality means that fewer births are required to reach a desired
number of surviving children.
Similarly, although short birth spacing generally leads to worse health outcomes, 
the more educated the mother, the more muted this relationship is, which, at least partly,
removes the health outcome incentive for longer birth spacing
\citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

Second, better-educated women are more efficient at producing child human capital 
\citep{Behrman1999}.
[need something on son preference? Maybe simply use increasing returns to male education]
Hence, families that value sons' education more will tend to both demand better-educated
women as mothers and have fewer children because both allow for higher quality boys.
With the increasing return to male education that India has experienced, this imply
an increased demand for better-educated women combined with lower fertility, and an 
increased use of sex selection when available.
In this case, the associations between higher female education, lower fertility, and more
sex selection are not a causal effect driven by higher wages but instead the result of the 
increased return to male education.

Even as the level of female education has increased, the female labor force 
participation in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
India's female labor force participation is now one of the lowest in the world  
and does not yet show any signs of increasing \citep{Klasen2015,Chatterjee2018}.
In line with previous research, the NFHS data show a U-shaped relationship between 
education and working for married women, with the highest percentage working being women 
with either no education or 12 or more years and the lowest being women with 8--11 years 
of education.%
\footnote{
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

@@
